Full-day, in home validation of infant body position measurements from inertial sensors

### 2023-03-01 manuscript.Rmd
current study draft

Accordingly, the overarching goal of the current study is to test the feasibility and validity of long-form body position recording in the home during unsupervised, everyday behavior. During a supervised recording, whether in the home or in the lab, the researcher can set up the situation to encourage or restrict certain behaviors and will usually instruct the caregiver to play with the infant. Within-day variations in activities (e.g., eating lunch in a high chair) create challenging situations for applying automated classification of body position. Will models trained on video recorded observations at the beginning of the day generalize to predict later behavior?

>>>>

Accordingly, the overarching goal of the current study is to test the feasibility and validity of long-form body position recording in the home during unsupervised, everyday behavior. During a supervised recording, whether in the home or in the lab, the researcher can set up the situation to encourage or restrict certain behaviors and will usually instruct the caregiver to play with the infant. Within-day variations in activities (e.g., eating lunch in a high chair) create challenging situations for applying automated classification of body position. Will models trained on video-recorded observations at the beginning of the day generalize to predict behavior at a later time?
Assessing the validity of temporally *distal* periods is a crucial step to establish whether automatic classification can be used to measure body position across a day.

In the current study, we report the feasibility and validity of body position classification over the full day in the home based on XX testing sessions from XX infants aged 4-14 months. Participants received a custom pair of infant leggings embedded with 4 IMUs (one on each ankle and one on each hip) and a video camera to collect ground truth data about infant body position. A *proximal comparison* period began when participants received the equipment and completed a guided phone call during which caregivers were asked to elicit different body positions based on prompts from the experimenter. Although not directly supervised, this period was most similar to previous recordings because it occurred during a convenient time for the infant and caregiver to play and receive instruction. The **first goal** of the current study was to determine the accuracy of body position classification during the proximal comparison period using our novel, semi-supervised procedure in participants' homes.

A second, *distal comparison* period followed the proximal comparison period and captured approximately 90 minutes of home behavior without any input from the experimenter. Following a long delay from the start of the study when caregivers were guided in placing the equipment, it is possible that accuracy could become worse if caregivers or infants moved the garment or sensors. Moreover, increasing variation in everyday activities creates a challenge for classification. Thus, the **second goal** of our study was to assess accuracy during the distal comparison.

After this second video recording period, we asked caregivers to have infants continue to wear IMUs for the rest of the day until they went to bed, creating the first real-time, full-day dataset of infant body position. Interpreting such data required caregivers to log when infants napped, when they removed the sensor garment for diaper changes or other reasons, and when infants went to bed at the end of the day. The **third goal** of the study was to examine the quality of the full-day data. Could body position be successfully recorded over the desired period? Finally, the novel full-day dataset affords us a unique chance to ask whether the estimated time infants spent in different body positions align with past results using video and survey methods. Thus, the **fourth goal** of the current study was to determine whether our full-day body position estimates yielded expected age-related differences in everyday body position. Based on past results, with age infants should spend increasingly more time sitting and upright but less time supine.

### 2023-03-01 manuscript.tex
current study draft

Measuring infant movement with inertial movement units (IMUs) is a promising avenue for long form recordings of motor behavior in the home. Lightweight sensors (XX-XXg) can be embedded in garments to make recordings fully \emph{mobile}, and they are \emph{unobtrusive} because they do not require a researcher's presence nor do they record sensitive audio/video that might influence participants' behavior. Many commerically-available and inexpensive IMUs have \textgreater{} 12 hour battery life an onboard storage to record \emph{real-time}, \emph{full-day} data of movements at a sampling rate that is higher than typical video (e.g., 50-100 Hz). Moreover, past work has successfully recorded the rate of leg kicks {[}SMITH CITE{]} and activity intensity {[}CITES{]} in infants and children across an entire day.

The open question is whether \emph{automatic classification} is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurement validity is acceptable over long recording periods. IMUs typically contain accelerometers that measure linear acceleration paired with gyroscopes that measure angular acceleration. Unlike motion tracking systems that might be used in a lab, IMUs do not provide data about the position of the body in space. Thus, data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration timeseries) into meaningful categories (e.g., supine, prone, sitting, upright). The difficulty of the classification task depends on the categories of interest. More basic aspects of movement, such as leg kicks or overall activity intensity, can be identified by taking the magnitude of acceleration (irrespective of direction) and applying thresholds or cut-points to define when a movement has occurred at a particular intensity.

Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is too complex for cut-point definitions to be accurate. First, the magnitude of movement can vary greatly \emph{within} a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant and be stationary in ``tummy time'' or crawling in myriad ways {[}ADOLPH{]}. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a ``V'' position with legs fully extended, in a ``W'' position with knees bent. Sitting on a caregivers' lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates still more possibilities for how the arms and legs may move relative to the torso. Indeed, creating an all-encompassing set of rules for how to annotate sitting from video is no trivial task because of the various ways that sitting can occur in daily life.

Thus, modern approaches to human activity recognition have used machine learning to classify activity categories based on features derived from IMU data in adults {[}CITEs{]}, children {[}CITEs{]}, and infants {[}CITEs{]}. Synchronized video with ground-truth human annotations creates training data for a machine learning algorithm, such as a random forest models, that can be later used to predict categories from future IMU data that was not annotated. Although in the current investigation we focus on classification of body position, this approach can be extended to categorize other aspects of movement, such as locomotion {[}CITE{]} and OTHER EXAMPLES? Our openly-shared dataset {[}CITE{]} contains synchronized video and IMU data, allowing researchers to annotate and train models for behaviors beyond body position.

REVIEW PAST WORK IN BODY POSITION CLASSIFICATION, END WITH WHAT IS MISSING.

@@

@@

>>>>

Measuring infant movement with inertial movement units (IMUs) is a promising avenue for long form recordings of motor behavior in the home. Lightweight sensors (XX-XXg) can be embedded in garments to make recordings fully \emph{mobile}, and they are \emph{unobtrusive} because they do not require a researcher's presence nor do they record sensitive audio/video that might influence participants' behavior {[}(Cliff et al., 2009; de Barbaro, 2019; Lobo et al., 2019; Bruijns et al., 2020){]}. Many commerically-available and inexpensive IMUs have \textgreater{} 12 hour battery life an onboard storage to record \emph{real-time}, \emph{full-day} data of movements at a sampling rate that is higher than typical video (e.g., 50-100 Hz). Moreover, past work has successfully recorded the rate of leg kicks {[}(Smith et al., 2017; Jiang et al., 2018){]} and activity intensity {[}Hagenbuchner et al., 2015; Trost et al., 2018{]} in infants and children across an entire day.

The open question is whether \emph{automatic classification} is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurement validity is acceptable over long recording periods. IMUs typically contain accelerometers that measure linear acceleration paired with gyroscopes that measure angular acceleration. Unlike motion tracking systems that might be used in a lab, IMUs do not provide data about the position of the body in space. Thus, data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration timeseries) into meaningful categories (e.g., supine, prone, sitting, upright). The difficulty of the classification task depends on the categories of interest. More basic aspects of movement, such as overall activity intensity, can be identified by taking the magnitude of acceleration (irrespective of direction) and applying thresholds or cut-points to define when a movement has occurred at a particular intensity {[}(Trost et al., 2012; Kuzik et al., 2015; Hager et al., 2017; Armstrong et al., 2019){]}.

Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is too complex for cut-point definitions to be accurate. First, the magnitude of movement can vary greatly \emph{within} a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant and be stationary in ``tummy time'' or crawling in myriad ways {[}ADOLPH{]}. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a ``V'' position with legs fully extended, in a ``W'' position with knees bent. Sitting on a caregivers' lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates still more possibilities for how the arms and legs may move relative to the torso. Indeed, creating an all-encompassing set of rules for how to annotate sitting from video is no trivial task because of the various ways that sitting can occur in daily life. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity {[}Kwon{]}.
GREENSPAN\ldots{}

Thus, modern approaches to human activity recognition have used machine learning to classify activity categories based on features derived from IMU data in adults {[}(Preece et al., 2009; Arif and Kattan, 2015){]}, children {[}(Nam and Park, 2013; Zhao et al., 2013; Ren et al., 2016; Stewart et al., 2018){]}, and infants {[}Franchak; Airaksinen; Yao{]}. Synchronized video with ground-truth human annotations creates training data for a machine learning algorithm, such as a random forest models, that can be later used to predict categories from future IMU data that was not annotated. Although in the current investigation we focus on classification of body position, this approach can be extended to categorize other aspects of movement, such as locomotion {[}CITE{]} and OTHER EXAMPLES? Accompanying our paper, we have created a database {[}CITE{]} containing synchronized video and IMU data, allowing researchers to annotate and train models for behaviors beyond body position.

Three prior investigations have used machine learning to categorize infant body position from IMUs towards the goal of collecting full-day data, and each has yielded encouraging validation results. Airaksinen et al. (2020) tested 4- to 8-month-olds in a laboratory visit, and found XX\% accuracy (kappa = XX\%) in distinguishing between body position categories that crawling infants could perform on the floor (excluding times that infants were held by caregivers). Using a wider age range of XX-XX, {[}FRANCHAK{]} found XX\% accuracy (kappa = XX\%) in a laboratory validation study in categorizing body position that included infants who could both crawl and walk and also categorized caregiver holding. Most recently, {[}Airaksinen NEW{]} conducted a validation study of body position classification in either a home or clinic testing in 4- to 19-month-olds, refining their previous method to detect moments that infants were carried by caregivers. Classification accuracy did not vary between lab and home settings, and was generally high (95\%, kappa = .93). However, in all three studies accuracy was assessed within a brief (15-60 minute) session supervised by a researcher, leaving the open question of how well body position classification will scale to testing across an entire day of natural home life. YAO?

@@

Accordingly, the overarching goal of the current study is to test the feasibility and validity of long-form body position recording in the home during unsupervised, everyday behavior. During a supervised recording, whether in the home or in the lab, the researcher can set up the situation to encourage or restrict certain behaviors and will usually instruct the caregiver to play with the infant. Within-day variations in activities (e.g., eating lunch in a high chair) create challenging situations for applying automated classification of body position. Will models trained on video-recorded observations at the beginning of the day generalize to predict behavior at a later time?
Assessing the validity of temporally \emph{distal} periods is a crucial step to establish whether automatic classification can be used to measure body position across a day.

In the current study, we report the feasibility and validity of body position classification over the full day in the home based on XX testing sessions from XX infants aged 4-14 months. Participants received a custom pair of infant leggings embedded with 4 IMUs (one on each ankle and one on each hip) and a video camera to collect ground truth data about infant body position. A \emph{proximal comparison} period began when participants received the equipment and completed a guided phone call during which caregivers were asked to elicit different body positions based on prompts from the experimenter. Although not directly supervised, this period was most similar to previous recordings because it occurred during a convenient time for the infant and caregiver to play and receive instruction. The \textbf{first goal} of the current study was to determine the accuracy of body position classification during the proximal comparison period using our novel, semi-supervised procedure in participants' homes.

A second, \emph{distal comparison} period followed the proximal comparison period and captured approximately 90 minutes of home behavior without any input from the experimenter. Following a long delay from the start of the study when caregivers were guided in placing the equipment, it is possible that accuracy could become worse if caregivers or infants moved the garment or sensors. Moreover, increasing variation in everyday activities creates a challenge for classification. Thus, the \textbf{second goal} of our study was to assess accuracy during the distal comparison.

After this second video recording period, we asked caregivers to have infants continue to wear IMUs for the rest of the day until they went to bed, creating the first real-time, full-day dataset of infant body position. Interpreting such data required caregivers to log when infants napped, when they removed the sensor garment for diaper changes or other reasons, and when infants went to bed at the end of the day. The \textbf{third goal} of the study was to examine the quality of the full-day data. Could body position be successfully recorded over the desired period? Finally, the novel full-day dataset affords us a unique chance to ask whether the estimated time infants spent in different body positions align with past results using video and survey methods. Thus, the \textbf{fourth goal} of the current study was to determine whether our full-day body position estimates yielded expected age-related differences in everyday body position. Based on past results, with age infants should spend increasingly more time sitting and upright but less time supine.

@@

\leavevmode\vadjust pre{\hypertarget{ref-AiraksinenRasanen2020}{}}%
Airaksinen, M., Räsänen, O., Ilén, E., Häyrinen, T., Kivi, A., Marchi, V., Gallen, A., Blom, S., Varhe, A., Kaartinen, N., et al. (2020). Automatic posture and movement tracking of infants with wearable movement sensors. \emph{Scientific Reports}, \emph{10}(1), 1--13.
